\section{Introduction}
\label{sec:introduction}

% ── ¶1  General landscape: AI for AMP is a thriving field ──
Antimicrobial resistance is among the most pressing global health
threats~\citep{who2019,oneill2016}, and antimicrobial peptides (AMPs)---short,
cationic, amphipathic sequences that disrupt bacterial membranes through
mechanisms different from conventional antibiotics---offer a promising class
of candidates for new therapeutics~\citep{zasloff2002,brogden2005,hancock2006}.
Discovering such peptides computationally has become a major research
direction.
The field now offers a rich and growing tool set:
supervised classifiers that screen sequences for antimicrobial
activity~\citep{amplify,ampscanner,iamppred,ampbert},
large protein language models whose embeddings transfer to peptide
tasks~\citep{esm2,prottrans,unirep},
lightweight peptide-specific language models that achieve competitive
performance at a fraction of the cost~\citep{pepbert,llamp,peptidebert},
and conditional generators that propose novel AMP
candidates~\citep{hydramp,diffamp,pandi2023,kppepgen,pepvae}.
These approaches have collectively advanced the state of computational AMP
discovery.

% ── ¶2  Observation: pre-trained representations increasingly central,
%        and the dominant pre-training objective is MLM ──
Across these approaches, learned sequence representations have become the
central building block---whether transferred from a general protein language
model or trained on a domain-specific peptide corpus.
The dominant pre-training objective for learning such representations is
masked language modeling (\mlmenc{}): mask out amino acids and predict the
exact identity of each masked token.
\mlmenc{} has been effective across protein science, and we do not argue that
it is wrong.
But it is worth noticing what this objective asks the model to learn:
accurate reconstruction of specific residues at specific positions.

% ── ¶3  Insight: AMP activity is a holistic pattern, not a token pattern ──
AMP activity, however, does not arise from individual residue identities.
What makes a peptide antimicrobial is a combination of emergent,
whole-sequence properties: sufficient net positive charge, an amphipathic
arrangement of hydrophobic and hydrophilic residues, appropriate length, and
the ability to adopt a membrane-active
conformation~\citep{fjell2012,hancock2006}.
The same functional role is often filled by different amino acids---replacing
one cationic residue with another may preserve activity, while a single
charge-removing mutation can abolish it.
In this sense, antimicrobial activity is better described as a
\emph{holistic physicochemical pattern} than as a sequence motif.
This raises a natural question: would a pre-training objective that operates
at the level of representations rather than tokens better capture how AMP
activity actually works?

% ── ¶4  Our proposal (tight: what we built and the core question) ──
We explore this question with \projectname{}, a compact (\encoderparams{}
parameter) encoder pre-trained on \corpussize{} AMP-like sequences using a
\jepafull{} (\jepa{})~\citep{jepa_assran}.
Unlike \mlmenc{}, \jepa{} predicts target \emph{representations} from context
representations in latent space, without reconstructing specific tokens.
We compare \projectname{} against \esmtwo{} and an identically architected
\mlmenc{} encoder to understand what changes when the pre-training signal
shifts from token reconstruction to representation prediction.

% ── ¶5+ RQs and contributions: TODO after wet-lab data ──
